---
title: "Assignment 1 Example: Survey Data"
author: "Carole Voulgaris"
date: "10/16/2021"
output: 
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction and setup

The purpose of this example is to demonstrate some of the methods that will be helpful to you in loading survey data (especially data from IPUMS) and calculating descriptive statistics.

I'll be using the following packages.

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(ipumsr)
library(survey)
library(srvyr)
```

# Research question

How does the amount of time a person spends traveling to work relate to their income, sex, mode of transportation, and hours worked per week?

# Prior research

Crane (2007) finds that there is a widening gap in commute times between men and women. Sandow and Westin (2010) find that men experience a higher income premium for long-distance commuting than women do.

# Data 

For this analysis, I'm using microdata from the 2019 5-year sample of the American Community Survey for the Los Angeles Metropolitan Area, as downloaded from IPUMS USA (Ruggles et al. 2021). I'm going to analyze the following five variables:

1. INCTOT: Annual personal income.
2. TRANTIME: Typical commute time to work
3. UHRSWORK: Typical hours worked per week
4. TRANWORK: Mode of transportation to work
5. SEX: Sex

```{r, message=FALSE, results='hide'}
ddi <- read_ipums_ddi("data/usa_00005.xml")
data <- read_ipums_micro(ddi) %>%
  na_if(9999999) %>%
  filter(UHRSWORK > 0) %>%
  mutate(SEX = as_factor(SEX)) %>%
  mutate(TRANWORK = lbl_collapse(TRANWORK, ~.val %/% 10)) %>%
  mutate(TRANWORK = as_factor(TRANWORK))
```
 
This dataset includes 348,564 individual survey respondents. 

The dataset includes weights to adjust for the ways in which the sample is not representative of the overall population, so I'll use the survey (Lumley, 2004) and srvyr (Freedman Ellis and Schneider 2021) packages to incorporate those weights into my analysis. First, I need to create a survey design object.

```{r}
svy_data <- data %>%
  as_survey_design(ids = CLUSTER,
                   strata = STRATA,
                   weights = PERWT)
```

# Descriptive statistics

## Continuous variables

Here's how I can create a table with the estimated mean and 95-percent confidence interval and variance (the square of the standard deviation) for each of my three continuous variables.

```{r}
travel_time_mean <- svy_data %>%
  summarize(mean = survey_mean(TRANTIME, vartype = c("ci", "var"), na.rm = TRUE))

income_mean <- svy_data %>%
  summarize(mean = survey_mean(INCTOT, vartype = c("ci", "var"), na.rm = TRUE))

hours_mean <- svy_data %>%
  summarize(mean = survey_mean(UHRSWORK, vartype = c("ci", "var"), na.rm = TRUE))
```

I can do something similar to create a table with the 25th, 50th, and 75th percentile values.

```{r}
travel_time_quants <- svy_data %>%
  summarize(quants = survey_quantile(TRANTIME, 
                                     quantiles =  c(0.25, 0.5, 0.75),
                                     vartype = NULL)) 

income_quants <- svy_data %>%
  summarize(quants = survey_quantile(INCTOT, 
                                     quantiles =  c(0.25, 0.5, 0.75),
                                     vartype = NULL)) 

hours_quants <- svy_data %>%
  summarize(quants = survey_quantile(UHRSWORK, 
                                     quantiles =  c(0.25, 0.5, 0.75),
                                     vartype = NULL)) 
```

I can combine those six tables into a single table using `rbind()` and `left_join()`, and I add the standard deviations and the interquartile ranges to that combined table.

```{r, message=FALSE}
mean_summary <- rbind(travel_time_mean, income_mean, hours_mean) %>%
    mutate(variable = c("Travel time to work",
                      "Personal income",
                      "Usual hours worked per week"))

quant_summary <- rbind(travel_time_quants, income_quants, hours_quants) %>%
    mutate(variable = c("Travel time to work",
                      "Personal income",
                      "Usual hours worked per week"))

summary <- left_join(mean_summary, quant_summary) %>%
  mutate(sd = mean_var^0.5) %>%
  mutate(IQR = quants_q75 - quants_q25) %>%
  rename(median = quants_q50) %>%
  select(variable, mean, mean_low, mean_upp, median, sd, IQR)

kable(summary, digits=2)
```

I can also create a histogram using `svyhist()`.

```{r}
travel_time_hist <- svyhist(~TRANTIME, design = svy_data,
                            freq = TRUE)
```

The above plot is pretty basic. I can also bring it into ggplot.

```{r}
travel_time_hist_df <- tibble(mids = travel_time_hist$mids,
                              counts = travel_time_hist$counts)

ggplot(travel_time_hist_df) +
  geom_bar(aes(x = mids, y = counts),
           stat = "identity") +
  theme_bw() +
  scale_x_continuous(name = "Travel time to work (minutes)") +
  scale_y_continuous(name = "Estimated number of workers (weighted)")
```

## Categorical variables

I can use `group_by()` and `survey_prop()` to calculate shares of my categorical variables, along with confidence intervals.

```{r}
ed_shares <- svy_data %>%
  group_by(SEX) %>%
  summarize(share = survey_prop(vartype = "ci"))

kable(ed_shares, digits = 3)
```

```{r}
mode_shares <- svy_data %>%
  group_by(TRANWORK) %>%
  summarize(share = survey_prop(vartype = "ci"))

kable(mode_shares, digits = 3)
```

# References

Crane, Randall. "Is there a quiet revolution in women's travel? Revisiting the gender gap in commuting." Journal of the American planning association 73, no. 3 (2007): 298-316.

Freedman Ellis, Greg, and Ben Schneider (2021). srvyr: 'dplyr'-Like Syntax for Summary Statistics of Survey Data. R package version 1.0.1. https://CRAN.R-project.org/package=srvyr

Lumley, T. (2004) Analysis of complex survey samples. Journal of Statistical Software 9(1): 1-19

Ruggles, Steven, Sarah Flood, Sophia Foster, Ronald Goeken, Jose Pacas, Megan Schouweiler and Matthew Sobek. IPUMS USA: Version 11.0 [dataset]. Minneapolis, MN: IPUMS (2021) https://doi.org/10.18128/D010.V11.0

Sandow, Erika, and Kerstin Westin. "The persevering commuter–Duration of long-distance commuting." Transportation Research Part A: Policy and Practice 44, no. 6 (2010): 433-445.