---
title: "Assignment 1 Example"
author: "Carole Voulgaris"
date: "10/16/2021"
output: 
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction and setup

I'll be using the following libraries in this example.

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(gridExtra)
```

The instructions for the assignment require you to summarize your data be presenting the following for each continuous variable:

1.	The sample mean
2.  The sample median
3.	The 95-percent confidence interval for the population mean
4.	The sample standard deviation
5.	The interquartile range
6.	An illustration (e.g. a histogram) and description of the distribution.

And the following for each categorical variable:
1.	The sample proportion in each category
2.	The 95-percent confidence interval for the population proportion in each category

I'm going to start by reading in a dataset on county-level COVID rates. The data includes six variables:

1. GEOID: An identifier for each county.
2. med_age_E: The estimated median age for county residents
3. pop_dens: The number of residents per square mile
4. covid_rate: The number of COVID cases per capita to data (October 15, 2021)
5. type: One of six county types (Large metro, large fringe metro, medium metro, small metro, micropolitan area, or non-core),
6. The party of the candidate receiving the most of the county's votes in the most recent presidential election.

```{r, message=FALSE}
data <- read_csv("covid_example.csv")
```

# Summarize continuous variables

The continuous variables I want to use are median age, population density, and COVID cases per capita.

## Calculations

I can get the sample mean and the 95-percent confidence interval from the population mean with the `t.test()` function (this is a one-sample t-test). 

```{r}
age_t_test <- t.test(data$med_age_E)
dens_t_test <- t.test(data$pop_dens)
covid_t_test <- t.test(data$covid_rate)
```

You might try typing `age_t_test` into your console to see what the output looks like, or you could expand that item in your environment tab to see what information it includes.

I can get the median and the interquartile range using the `quantile()` function.

```{r}
age_quartiles <- quantile(data$med_age_E, na.rm = TRUE)
dens_quartiles <- quantile(data$pop_dens, na.rm = TRUE)
covid_quartiles <- quantile(data$covid_rate, na.rm = TRUE)
```

Again, you could try typing `dens_quartiles` into your console to see what the output looks like.

And I can get the standard deviation from using the `sd()` function. 

```{r}
age_st_dev <- sd(data$med_age_E, na.rm = TRUE)
dens_st_dev <- sd(data$pop_dens, na.rm = TRUE)
covid_st_dev <- sd(data$covid_rate, na.rm = TRUE)

```

You might try typing covid_st_dev into your console to see the value.

I can use the geom_histogram() function within a ggplot object to generate a histogram for each variable. I'm going to log-transform the x-axis for the population density so the variation is easier to see.

```{r, message=FALSE, warning=FALSE}
age_hist <- ggplot(data) +
  geom_histogram(aes(x = med_age_E),
                 bins = 30)

dens_hist <- ggplot(data) +
  geom_histogram(aes(x = pop_dens),
                 bins = 30) +
  scale_x_continuous(trans = "log")

covid_hist <- ggplot(data) +
  geom_histogram(aes(x = covid_rate),
                 bins = 30)
```

If you type age_hist, dens_hist, or covid_hist into your console, the plots will display on your plot tab. Try recreating dens_hist without the log transformation to see why the log transformation was advisable.

## Displaying your results

### Minimal formatting

You have some options for how to display your results. You could type names of the summary variables you just created into a code chunk to print out the results. For example, here are all the summary statistics for COVID rates, followed by the histogram for that variable:

```{r, warning = FALSE}
covid_t_test
covid_quartiles
covid_st_dev
covid_hist
```

### Formatted table

You could also create a nice table with all your results. Here is how you might do that.

```{r}
cont_summary <- tibble(
  Variable = c("Median age", 
               "Population density (people per square mile)", 
               "COVID rate (Total confirmed cases per 1,000 residents)"),
  `Sample mean` = c(age_t_test$estimate,
                    dens_t_test$estimate,
                    covid_t_test$estimate * 1000),
  `Population mean (95% confidence) - low` = 
    c(age_t_test$conf.int[1],
      dens_t_test$conf.int[1],
      covid_t_test$conf.int[1] * 1000),
  `Population mean (95% confidence) - high` =
    c(age_t_test$conf.int[2],
      dens_t_test$conf.int[2],
      covid_t_test$conf.int[2] * 1000),
  Median = c(age_quartiles[3],
             dens_quartiles[3],
             covid_quartiles[3] * 1000),
  `Interquartile range` = c(age_quartiles[4] - age_quartiles[2],
                            dens_quartiles[4] - dens_quartiles[2],
                            (covid_quartiles[4] - covid_quartiles[2]) * 1000),
  `Standard deviation` = c(age_st_dev,
                          dens_st_dev,
                          covid_st_dev * 1000))

kable(cont_summary, digits = 0)
```

### Row of histograms

And then you could display the three histograms in a neat little row (I'm setting `fig.height=3` and `fig.width=7` in the heading of the code chunk).

```{r, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
pretty_age_hist <- age_hist +
  theme_bw() +
  scale_x_continuous(name = "Median age") +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90))

pretty_dens_hist <- dens_hist +
  theme_bw() + 
  scale_x_continuous(name = "Population density\n(residents per square mile)",
                     trans = "log",
                     breaks = c(0.1, 1, 10, 100, 1000, 10000),
                     labels = c("0.1", "1", "10", "100", "1000", "10,000")) +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90))

pretty_covid_hist = covid_hist +
  theme_bw() +
  scale_x_continuous(name = "Covid cases per\n1,000 residents",
                     breaks = seq(0, 0.3, by=.05),
                     labels = seq(0, 300, by = 50)) +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90)) 

grid.arrange(pretty_age_hist, pretty_dens_hist, pretty_covid_hist,
             ncol = 3)
```

# Summarizing categorical variables

I have two categorical variables:

1. The party of the candidate the won the most votes in the 2020 presidential election and
2. The county type, based on the NCHS classification.

## Calculations

I can use a one sample t-test to get the 95-percent confidence interval for the proportion of the sample in each category

```{r}
pct_repub <- t.test(data$majority_vote == "Republican")
pct_dem <-  t.test(data$majority_vote == "Democrat")

pct_large_metro <- t.test(data$type == "Large central metro")
pct_fringe_metro <- t.test(data$type == "Large fringe metro")
pct_med_metro <- t.test(data$type == "Medium metro")
pct_small_metro <- t.test(data$type == "Small metro")
pct_micro <- t.test(data$type == "Micropolitan")
pct_rural <- t.test(data$type == "Non-core")
```

## Displaying your results

Now I can make some choices about how to display my results

### Formatted tables

I could create a nice formatted table using the same approach I did for the categorical variables.

```{r}
cat_summary_party <- tibble(`Majority Party in 2020` = 
                              c("Republican",
                                "Democrat"),
                            `Sample proportion` = 
                              c(pct_repub$estimate * 100,
                                pct_dem$estimate *100),
                            `95-percent confidence - low` = 
                              c(pct_repub$conf.int[1] * 100,
                                pct_dem$conf.int[1] * 100),
                            `95-percent confidence - high` = 
                              c(pct_repub$conf.int[2] * 100,
                                pct_dem$conf.int[2] * 100))

kable(cat_summary_party, digits = 0)
```


```{r}
cat_summary_type <- tibble(`County type` = 
                              c("Larch central metro",
                                "Large fringe metro",
                                "Medium metro",
                                "Small metro",
                                "Micropolitan",
                                "Non-core"),
                            `Sample proportion` = 
                              c(pct_large_metro$estimate * 100,
                                pct_fringe_metro$estimate * 100,
                                pct_med_metro$estimate * 100,
                                pct_small_metro$estimate * 100,
                                pct_micro$estimate * 100,
                                pct_rural$estimate * 100),
                            `95-percent confidence - low` = 
                              c(pct_large_metro$conf.int[1] * 100,
                                pct_fringe_metro$conf.int[1] * 100,
                                pct_med_metro$conf.int[1] * 100,
                                pct_small_metro$conf.int[1] * 100,
                                pct_micro$conf.int[1] * 100,
                                pct_rural$conf.int[1] * 100),
                            `95-percent confidence - high` = 
                              c(pct_large_metro$conf.int[2] * 100,
                                pct_fringe_metro$conf.int[2] * 100,
                                pct_med_metro$conf.int[2] * 100,
                                pct_small_metro$conf.int[2] * 100,
                                pct_micro$conf.int[2] * 100,
                                pct_rural$conf.int[2] * 100))

kable(cat_summary_type, digits = 0)
```

### Bar charts

It could also be helpful to visualize these results as a bar chart with error bars to indicate the confidence intervals. Here is a bar chart for the majority vote in the past presidential election.

```{r}
ggplot(cat_summary_party) +
  geom_bar(aes(x = `Majority Party in 2020`, 
               y = `Sample proportion`),
           stat = "identity") +
  geom_errorbar(aes(x = `Majority Party in 2020`, 
               y = `Sample proportion`,
               ymin = `95-percent confidence - low`,
               ymax = `95-percent confidence - high`),
           stat = "identity") +
  scale_y_continuous(name = "Percent of counties",
                     breaks = c(0, 20, 40, 60, 80),
                     labels = c("0", "20%", "40%", "60%", "80%")) +
  theme_bw()
```

And here is one for the county type.

```{r}
ggplot(cat_summary_type) +
  geom_bar(aes(x = `County type`, 
               y = `Sample proportion`),
           stat = "identity") +
  geom_errorbar(aes(x = `County type`, 
               y = `Sample proportion`,
               ymin = `95-percent confidence - low`,
               ymax = `95-percent confidence - high`),
           stat = "identity") +
  scale_y_continuous(name = "Percent of counties",
                     breaks = c(0, 10, 20, 30, 40),
                     labels = c("0", "10%", "20%", "30%", "40%")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) 
```