---
title: "Data Assembly"
output: 
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidycensus)
```


# Introduction

The purpose of this exercise is to determine the relationship between the umemployment rate in a region and each of four other variables:

* Dominant industry
* Metropolitan or micropolitan area
* Mean usual hours worked
* Median household income

The following code assembles the dataset (to run the code, you'll need to have a census API key installed on your computer).

```{r, message=FALSE}
variables <- c(num_unemp_16up_ = "B23025_005",
               num_workers_16up_ = "B23025_002",
               mean_work_hours_ = "B23020_001",
               median_income_ = 	"B06011_001",
               tot_wrkrs_ind_ = "B08126_001",
               wrkrs_ag_ = "B08126_002",
               wrkrs_const_ = "B08126_003",
               wrkrs_manuf_ = "B08126_004",
               wrkrs_whlsl_ = "B08126_005",
               wrkrs_retail_ = "B08126_006",
               wrkrs_transpo_ = "B08126_007",
               wrkrs_info_ = "B08126_008",
               wrkrs_FIRE_ = "B08126_009",
               wrkrs_prof_ = "B08126_010",
               wrkrs_edu_ = "B08126_011",
               wrkrs_rec_ = "B08126_012",
               wrkrs_srvc_ = "B08126_013",
               wrkrs_PA_ = "B08126_014",
               wrkrs_mil_ = "B08126_015")

cbsa_data <- get_acs(geography = "cbsa",
                     variables = variables,
                     output = "wide") 

cbsa_data <- cbsa_data %>%
  mutate(main_ind_x = max.col(cbsa_data[,13:40])) %>%
  mutate(main_ind = case_when(main_ind_x == 1 ~ "Agriculture",
                              main_ind_x == 5 ~ "Manufacturing",
                              main_ind_x == 9 ~ "Retail sales",
                              main_ind_x == 17 ~ "Professional services",
                              main_ind_x == 19 ~ "Educational services",
                              main_ind_x == 21 ~ "Hospitality",
                              main_ind_x == 25 ~ "Public administration",
                              main_ind_x == 27 ~ "Military")) %>%
  mutate(metro = ifelse(str_detect(NAME, "Metro"), "Metropolitan", "Micropolitan")) %>%
  mutate(unemp_rate = num_unemp_16up_E / num_workers_16up_E) %>%
  rename(mean_work_hours = mean_work_hours_E) %>%
  rename(median_income = median_income_E) %>%
  select(GEOID, NAME, unemp_rate, median_income, mean_work_hours, metro, main_ind)
```

# Bivariate analysis

Calculate the magnitude and the significance of the relationship. Include a few sentences to describe each relationship.

## Relationship between unemployment rate and dominant industry

## Relationship between unemployment rate and metropolitan area status

## Relationship between unemployment rate and median income

## Relationship between unemployment rate and mean unsual hours worked
